Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-11-28 analysis/paper/paper.qmd
Wrote conclusion

abstract: |
 The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterizes the transition from the Early to Late Paleolithic in Korea. Stemmed points have been considered a representative tool that led this set of changes. In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic. Our main questions are: What were stemmed points used for? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point use? We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears. We analyzed TCSA with other variables including raw materials, weight, site, radiocarbon dates and locations. Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. TCSA values are controlled mostly by size rather than raw material types. We found different usage of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment. Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but SP were mainly used before Last Glacial Maximum (LGM). We observed that SP was only located in certain ecoregions in Korea but no clear functional pattern is confirmed. We conclude that stemmed points were multi-functional tools, with many likely designed for javelin and stabbing spear tips.

@@

Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabalistically discriminate different between functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. It is important to note that the TCSA metric alone cannot unambiguously determine artifact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa. A key advantage of the TCSA metric is its convenience of application: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate metric, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points. The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip]. @sisk2011african also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

@@

To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, and artifact size. To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points. We distinguished the research time period into three based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and examined the TCSA pattern. We explored the relationship between TCSA values and the location of assemblages by applying eco-regional information. We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges). Among our dataset, we excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data. For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites. Each analysis uses slightly different number of artifacts due to data availability.

@@

@@

We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values. Among `r n_sites` artifacts, we combined sites that contains fewer than 5 stemmed points aside and named them as "Other". We also explored possible environmental impact by connecting stemmed points to the eco-regional zones. We chose two maps of modern eco-regional zones and modified them for this research. @yim1975distribution first created vegetation forest map defined temperature and precipitations. This map then modified by @yi2010vegetation. They divided South Korea, our research area, into three zones: Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW). We also used a map from @lee2008relationship, which separated our research area into 16 different zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation. We only focused on 14 zones, excluding two islands that have no records of Paleolithic material culture: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone. It is acknowledged that modern eco-regional maps may not be suitable for studying the Paleolithic period directly, but they are the only maps of eco-regional seperation that are currently available to us.

@@

We computed artifact TCSA values for `r n_sites` sites that contain more than 5 stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than 5 stemmed points were grouped under the category of 'other'. Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values. This suggests that people made stemmed points for a variety of different purposes in Suyanggae. The TCSA values for Nosanri, Sibuk, and Wolpyeng are have a narrower range of lower values, which indicates less diverse functions for stemmed points. Yongsujaeul shows a narrower range but higher TCSA value. Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.

```{r}
#| label: fig-gis-tcsa-sites
#| fig-cap: "TCSA values by assemblages and ecoregions. A. Modern vegetation map defined by temporature and precipitation (modified the map from @yi2010vegetation, but original source is @yim1975distribution). The individual points indicate assemblages. B. Boxplots for  vegitation zones. The individual points indicate artifacts. C. Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship). D. Boxplots for  ecoregion zones, excludings zones with no artifacts."
#| fig-width:   5
#| fig-height:  4

@@

```{r}
#| label: fig-tukey-region
#| fig-cap: "."
#| fig-width:   5
#| fig-height:  4

#vegetation region#
points_in_veg_zones_all$ecoregion = as.factor(points_in_veg_zones_all$ecoregion)
eco_aov <- aov(min_tcsa ~ ecoregion, points_in_veg_zones_all)

@@

plot_grid(eco_plot, 
          for_plot, 
          eco_box, 
          for_box, 
          eco_tur_box, 
          for_tur_box,
          labels = c('A', 'C', 'B', 'D', 'E', 'F'),
          ncol = 2)

@@

In order to explore spatial patterns as well as the relationship between the function of the tools and the environment, we located individual assemblages on maps that depict eco-regional boundaries. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites) of eco-region map. We observed that TCSA values are strongely associated with the zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). To explore the TCSA distribution within the eco-regional zones, we excluded zones in the ecoregion map that have no stemmed points (@fig-gis-tcsa-sites panel D). We found no clear functional pattern throughout the zones, showing a wide variation in TCSA regardless of the density of stemmed point in each zone (@fig-gis-tcsa-sites panel B and D).

We then computed Tukey's for comparison tests between the eco-regional zones. Our results show that ST and CT, and SWT-ST have significantly different while SWT and CT are close to each other (@fig-tukey-region panel A). We found no significant difference within the zones of eco-region map (@fig-tukey-region panel B).

@@

To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely artifact uses by comparing the results to other archaeological and ethnographic studies Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point uses?

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). But in general, we found a wide range of TCSA in our sample, indicating that stemmed points played a diverse role in foraging toolkits. TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size). Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon). But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon-box). We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in function (@fig-gis-tcsa-sites). Both temporal and spatial analyses show that stemmed points were used for a particular environment, not so much harsh as LGM. @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points. Our results support this claim by showing the dense distribution of stemmed point in certain environment.

Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances. To examine the possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did. The simulated temperature distribution in south Korea during the Late Paleolithic period (Figure 9: A from @park2022did) seems similar to the modern eco-regional zones that we used in this research (@fig-gis-tcsa-sites). Figure 9: C from @park2022did shows the slow decrease of temperature until the end of LGM, which could explain our results of temporal pattern, meaning stemmed points were used during the cooler environment with a temperature between 4 and 6 ℃ (@fig-tcsa-radio-carbon-box). Results of SPD (Figure 10 from @park2022did) show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline during our research period, indicating no major impact from population dynamics on stemmed points.

@@

RESTATE QUESTIONS AND ANSWERS VERY BRIELFY

possible role
indicate people used stemmed poitns in different way
examined their likely uses

SUGGEST SOME POSSIBLE DIRECTIONS FOR FUTURE WORK, E.G. EXPERIMENTAL ARCHAEOLOGY, USEWEAR ANALYSIS

We are aware of the arbitrary that discriminating the function of small projectile points could be arbitrary work [@erlandson2014darts]. Because TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increased or decrease by shape of tip and cross section), we consider it a useful metric for discriminating between weapon tip functions [@hughes1998getting; @sitton2020tip; @lombard2021variation]. Stemmed points have been considered so far as long-distance hunting amateurs pursuing large bodies of prey such as a deer [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human]. TCSA enabled us to gain a more detailed understanding of the function of stemmed points.

>>>>

abstract: |
 The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterizes the transition from the Early to Late Paleolithic in Korea. Stemmed points have been considered a representative tool that led this set of changes. In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic. Our main questions are: What were stemmed points used for? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point use? We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears. We analyzed TCSA with other variables including raw materials, weight, radiocarbon dates and locations. Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. TCSA values are controlled mostly by size rather than raw material types. We found different usage of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment. Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but SP were mainly used before Last Glacial Maximum (LGM). We observed that SP was only located in certain ecoregions in Korea but no clear functional pattern is confirmed. We conclude that stemmed points were multi-functional tools, with many likely designed for javelin and stabbing spear tips.

@@

Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabalistically discriminate difference between functional classes of projectile armatures such as spearthrower (a.k.a. atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins]. It is important to note that the TCSA metric alone cannot unambiguously determine artifact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips. The TCSA metric represents the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation]. This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use. @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa. A key advantage of the TCSA metric is its convenience of application: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip]. Later, @sisk2011african proposed an alternate metric, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect. However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points. The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip]. @sisk2011african also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.

@@

To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, and artifact size. To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points. We distinguished the research time period into three based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and examined the TCSA pattern. We explored the relationship between TCSA values and the location of assemblages by applying eco-regional information.

@@

We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges). Among our dataset, we excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data. For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites. Each analysis uses slightly different number of artifacts due to data availability.

@@

We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values. Among `r n_sites` artifacts, we combined sites that contains fewer than 5 stemmed points aside and named them as "Other". We also explored possible environmental impact by connecting stemmed points to the eco-regional zones. We chose two maps of modern eco-regional zones and modified them for this research. @yim1975distribution first created vegetation forest map defined temperature and precipitations. This map then modified by @yi2010vegetation. They divided South Korea, our research area, into three zones: Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SWT). We also used a map from @lee2008relationship, which separated our research area into 16 different zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation. We only focused on 14 zones, excluding two islands that have no records of Paleolithic material culture: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone. It is acknowledged that modern eco-regional maps may not be suitable for studying the Paleolithic period directly, but they are the only maps of eco-regional seperation that are currently available to us.

@@

We computed artifact TCSA values for `r n_sites` sites that contain more than 5 stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than 5 stemmed points were grouped under the category of "Other". Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values. This suggests that people made stemmed points for accomplishing a variety of tasks at Suyanggae site. Nosanri, Sibuk, and Wolpyeng have a narrower range of lower values, which indicates less diverse likely uses for stemmed points. Yongsujaeul shows a narrower range but higher TCSA value. Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.

```{r}
#| label: fig-gis-tcsa-sites
#| fig-cap: "TCSA values by assemblages and ecoregions. A: Modern vegetation map defined by temporature and precipitation. The individual points indicate assemblages. B: TCSA distribution for vegitation zones. The individual points indicate artifacts. C: Tukey test result for vegitation zones. D: Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship). E: TCSA distribution for ecoregion zones, excluding zones with no artifacts. F: Tukey test result for ecoregion zones"
#| fig-width:   4.3
#| fig-height:  5.8

@@

######Tukey test######
#vegetation region#
points_in_veg_zones_all$ecoregion = as.factor(points_in_veg_zones_all$ecoregion)
eco_aov <- aov(min_tcsa ~ ecoregion, points_in_veg_zones_all)

@@

plot_grid(eco_plot, 
          for_plot, 
          eco_box, 
          for_box, 
          eco_tur_box, 
          for_tur_box,
          labels = c('A', 'D', 'B', 'E', 'C', 'F'),
          ncol = 2)

@@

In order to explore spatial patterns as well as the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional boundaries. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map. We observed that TCSA values are strongely associated with the zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). To explore the TCSA distribution within the eco-regional zones, we excluded zones in the ecoregion map that have no stemmed points (@fig-gis-tcsa-sites panel E). We found no clear distribution pattern throughout the zones, showing a wide variation in TCSA regardless of the density of stemmed point in each zone. For example, explored the six artifacts from South Temperature Zone (ST) to examine the gap of TCSA values among the artifacts (i.e. Three artifacts are over 175 while the other three are lower around or lower than 100. @fig-gis-tcsa-sites panel B). We observed two artifacts from Bonggok site with no age difference, both made out of rhyolite, have huge difference in TCSA values, indicating no pattern for TCSA variation.

We then computed Tukey's for comparison tests between the eco-regional zones. Our results show that ST and CT, and Subtropical-warm Temperature Zone (SWT)-ST have significantly different while SWT and CT are close to each other (@fig-gis-tcsa-sites panel C). However, we found no significant difference within the zones of eco-region map (@fig-gis-tcsa-sites panel F).

@@

To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely uses by comparing the results to the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges). Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point uses?

Our results show that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). But in general, we found a wide range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits. TCSA can be impacted by other factors such as raw materials and portability. @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size). Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A). But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon panel B). We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in function (@fig-gis-tcsa-sites). Both temporal and spatial analyses show that stemmed points were made for a particular environment, not so much harsh as LGM. @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points. Our results support this claim by showing the dense distribution of stemmed point in certain environment.

Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances. To examine the possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did. The simulated temperature distribution in south Korea during the Late Paleolithic period (Figure 9: A from @park2022did) seems similar to the modern eco-regional zones that we used in this research (@fig-gis-tcsa-sites panel A & D). Figure 9: C from @park2022did shows the slow decrease of temperature until the end of LGM, which could explain our results of temporal pattern, meaning stemmed points were used during the cooler environment with a temperature between 4 and 6 ℃. Results of SPD (Figure 10 from @park2022did) show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline during our research period, indicating no major impact from population dynamics on stemmed points.

@@

Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses of stemmed points by asking three research questions: What were stemmed points used for? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point use? We applied tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e width and thickness) to compute the metric. We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites. Then we examined the TCSA values with other variables including raw materials, weight, radiocarbon dates, and site locations. Based on the evolutionary approach, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed. We examined the possible impact of LGM on stemmed points and their distribution in eco-regional maps.

According to the different weapon-delivery systems that summarize TCSA of projectile points from archaeological and ethnographic cases, a majority of stemmed points might be used the most likely as javelin tips and stabbing spear tips. But overall we observed a wide range and also difference distribution of TCSA in each assemblage. Therefore, we conclude that stemmed points might play diverse roles during the Late Paleolithic. Based on our results, we assume that people were exposed to largely unfamiliar situation and created a stemmed point as multifunctional tool to performed multiple tasks. We are aware that discriminating the likely use of small projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increased or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation]. In our future research, we will conduct usewear analysis on stemmed points and experimental research as another attempt to identify detailed function of stemmed points and the technological connection between hunter-gatherer groups.
